Supplementary material for ``Screening Sinkhorn Algorithm for regularized Optimal Transport''

### 2019-05-16 tex/v3/section_1_introduction.tex
new

Our focus in this paper is a new analysis of Sinkhorn algorithm for approximating regularized OT distance between discrete measures.
The proposed approach is based on reformulating the dual of Sinkhorn divergence by adding new constraints feasibility.
These constraints are defined through a convex set which depends on two parameters acting like threshold and scaling factor.
We prove that dual solution of this reformulation guarantees the existence of two active indices sets for the potential variables.
Furthermore, the active sets are both directly linked to a priori fixed number budget of points from the supports of the given discrete measures.
We then restrict the constraints feasibility by taking into account the properties providing by the active sets to get a ``screened'' version of the dual of Sinkhorn divergence. 
Screenkhorn algorithm developed in this paper consists of two steps; the first one is an initialization step devoted to determine the active sets while the second is a constrained L-BFGS-B solver, a limited-memory quasi-Newton algorithm~\cite{nocedal1980,byrd1995L-BFGS-B} that relies on an estimation of the inverse of the Hessian based on gradients differences. 
L-BFGS-B algorithm was adapted to the OT setting via the dual of Sinkhorn divergence in ~\cite{cuturi2016SIAM,blondel2018ICML}.

>>>>

Our focus in this paper is a new analysis of Sinkhorn algorithm for approximating regularized OT distance between discrete measures.
The proposed approach is based on reformulating the dual of Sinkhorn divergence by adding new constraints feasibility.
These constraints are defined through a convex set which depends on two parameters acting like threshold and scaling factor.
We prove that dual solution of this reformulation guarantees the existence of two active indices sets for the potential variables.
Furthermore, the active sets are both directly linked to a priori fixed number budget of points from the supports of the given discrete measures.
We then restrict the constraints feasibility by taking into account the properties providing by the active sets to get a ``screened'' version of the dual of Sinkhorn divergence. 
Screenkhorn algorithm developed in this paper consists of two steps; the first one is an initialization step devoted to determine the active sets while the second is a constrained box-constrained L-BFGS-B solver, a limited-memory quasi-Newton algorithm~\cite{nocedal1980,byrd1995L-BFGS-B} that relies on an estimation of the inverse of the Hessian based on gradients differences. 
L-BFGS-B algorithm was adapted to the OT setting via the dual of Sinkhorn divergence in ~\cite{cuturi2016SIAM,blondel2018ICML}.

### 2019-05-16 tex/v3/section_3_algorithm.tex
new

The key idea of our approach is motivated by the so-called \emph{static screening test}~\cite{Ghaoui2010SafeFE} in supervised learning, that is a method able to quickly identify inactive features, i.e., features that have zero components in the solution vector. 
Then, these inactive features can be removed from the optimization problem to reduce its scale.
Before diving into detailed algorithmic analysis, let's us present a brief illustration of how we adapt static screening test to the dual of Sinkhorn divergence.
Towards this end, we define the convex set $\mathcal{C}^r_{\alpha} \subseteq \R^r$, for $r\in \mathbb N$ and $\alpha >0$, by $\mathcal{C}^r_{\alpha} = \{w \in \R^{r}: \min_{1\leq i\leq r} e^{w_i} \geq \alpha\}$.
In Figure~\ref{fig:motivations}, we plot $(e^{u^\star}, e^{v^\star})$ solution of dual of Sinkhorn divergence~\ref{sinkhorn-dual} in the particular case of: $n=m=500, \eta=0.1, \mu = \nu = \frac 1n \mathbf 1_n$ and the cost matrix $C$ correspond the pairwise euclidean distance, i.e., $C_{ij} = \norm{x_i - y_j}_2$. 
We also plot two lines correspond to $e^{u^\star} \equiv \alpha_u$ and $e^{v^\star} \equiv \alpha_v$ for some $\alpha_u>0$ and $\alpha_v >0$, playing the role of thresholds to select indices to be discard from the optimization procedure.

\paragraph{Static screening test.} The core of the static screening test developed in this work aims at locating two subsets of indices $(I, J)$ in $\{1, \ldots, n\}\times\{1, \ldots, m\}$ satisfying: $e^{u_i}\geq \alpha_u, \text{ and } e^{v_j}\geq \alpha_v, \text{ for all } (i,j) \in I \times J$ and 
$e^{u_{i'}} = \alpha_u, \text{ and } e^{v_{j'}} = \alpha_v, \text{ for all } (i',j') \in I^\complement \times J^\complement$, namely $(u,v) \in \mathcal{C}^n_{\alpha_u}\times \mathcal{C}^m_{\alpha_v}$.
The sets $I$ and $J$ will be called active sets, we further emphasize that they are determined in an initialization step in the proposed algorithm before running the L-BFGS-B solver.
Consequently, we can work on a "screened" dual of Sinkhorn divergence by considering variables that belong only to $\R^{|I|} \times \R^{|J|}$ in order to reduce the computational cost.

In the following, the parameters $\alpha_u = \varepsilon \kappa^{-1}$ and $\alpha_v = \varepsilon \kappa$ where $\varepsilon > 0$ and $\kappa > 0$ are fixed numeric constants to be given later. 
Our screening test relies on a new formulation of problem~\eqref{sinkhorn-dual}, that we refer as to \emph{approximate dual of Sinkhorn divergence}, given by:

\begin{equation} 
\label{screen-sinkhorn}
\min_{u \in \mathcal{C}^n_{\frac \varepsilon \kappa}, v\in \mathcal{C}^m_{\varepsilon\kappa}} \big\{\Psi_{\kappa}(u,v):= \mathbf{1}_n^\top B(u,v)\mathbf{1}_m - \inr{\kappa u, \mu} - \inr{\frac v\kappa, \nu} \big\},
\end{equation}
The $\kappa$-parameter plays a role of scaling factor, allowing to get a closed order of the potential variables $e^u$ and $e^v$, while the $\varepsilon$-parameter acts like a threshold for these variables.
Note that the approximate dual of Sinkhorn divergence coincides with the dual of Sinkhorn divergence~\eqref{sinkhorn-dual} in the setting of $\varepsilon=0$ and $\kappa=1$.
As explained above, our static screening test is based on constructing two {active sets} denoted $I_{\varepsilon, \kappa}$ and $J_{\varepsilon, \kappa}$ throughout the dual problem of~\eqref{screen-sinkhorn} as follows:

>>>>

The key idea of our approach is motivated by the so-called \emph{static screening test}~\cite{Ghaoui2010SafeFE} in supervised learning, which is a method able to quickly identify inactive features, i.e., features that have zero components in the solution vector. 
Then, these inactive features can be removed from the optimization problem to reduce its scale.
Before diving into detailed algorithmic analysis, let's us present a brief illustration of how we adapt static screening test to the dual of Sinkhorn divergence.
Towards this end, we define the convex set $\mathcal{C}^r_{\alpha} \subseteq \R^r$, for $r\in \mathbb N$ and $\alpha >0$, by $\mathcal{C}^r_{\alpha} = \{w \in \R^{r}: \min_{1\leq i\leq r} e^{w_i} \geq \alpha\}$.
In Figure~\ref{fig:motivations}, we plot $(e^{u^\star}, e^{v^\star})$ solution of dual of Sinkhorn divergence~\ref{sinkhorn-dual} in the particular case of: $n=m=500, \eta=0.1, \mu = \nu = \frac 1n \mathbf 1_n$ and the cost matrix $C$ correspond the pairwise euclidean distance, i.e., $C_{ij} = \norm{x_i - y_j}_2$. 
We also plot two lines correspond to $e^{u^\star} \equiv \alpha_u$ and $e^{v^\star} \equiv \alpha_v$ for some $\alpha_u>0$ and $\alpha_v >0$, playing the role of thresholds to select indices to be discard from the optimization procedure.

\paragraph{Static screening test.} The core of the static screening test developed in this work aims at locating two subsets of indices $(I, J)$ in $\{1, \ldots, n\}\times\{1, \ldots, m\}$ satisfying: $e^{u_i}\geq \alpha_u, \text{ and } e^{v_j}\geq \alpha_v, \text{ for all } (i,j) \in I \times J$ and 
$e^{u_{i'}} = \alpha_u, \text{ and } e^{v_{j'}} = \alpha_v, \text{ for all } (i',j') \in I^\complement \times J^\complement$, namely $(u,v) \in \mathcal{C}^n_{\alpha_u}\times \mathcal{C}^m_{\alpha_v}$.
The sets $I$ and $J$ are called active sets, we further emphasize that they are determined in an initialization step in the proposed algorithm before running the L-BFGS-B solver.
Consequently, we can work on a "screened" dual of Sinkhorn divergence by considering variables that belong only to $\R^{|I|} \times \R^{|J|}$ in order to reduce the computational cost.

In the following, the parameters $\alpha_u = \varepsilon \kappa^{-1}$ and $\alpha_v = \varepsilon \kappa$ where $\varepsilon > 0$ and $\kappa > 0$ are fixed numeric constants to be given later. 
Our screening test relies on a new formulation of problem~\eqref{sinkhorn-dual}, that we refer as to \emph{approximate dual of Sinkhorn divergence}, defined by:

\begin{equation} 
\label{screen-sinkhorn}
\min_{u \in \mathcal{C}^n_{\frac \varepsilon \kappa}, v\in \mathcal{C}^m_{\varepsilon\kappa}} \big\{\Psi_{\kappa}(u,v):= \mathbf{1}_n^\top B(u,v)\mathbf{1}_m - \inr{\kappa u, \mu} - \inr{\frac v\kappa, \nu} \big\},
\end{equation}
The $\kappa$-parameter plays a role of scaling factor, allowing to get a closed order of the potential components $e^u$ and $e^v$, while the $\varepsilon$-parameter acts like a threshold for these components.
Note that the approximate dual of Sinkhorn divergence coincides with the dual of Sinkhorn divergence~\eqref{sinkhorn-dual} in the setting of $\varepsilon=0$ and $\kappa=1$.
As explained above, our static screening test is based on constructing two {active sets} denoted $I_{\varepsilon, \kappa}$ and $J_{\varepsilon, \kappa}$ throughout the dual problem of~\eqref{screen-sinkhorn} as follows:

### 2019-05-16 tex/v3/section_4_marginal_viols.tex
new

This section is devoted to establishing theoretical guarantees for the marginal violations of Screenkhorn. 
We first define the screened marginals $\mu^{\text{sc}} = B(u^{\text{sc}}, v^{\text{sc}}) \mathbf 1_m$ and $\nu^{\text{sc}} = B(u^{\text{sc}}, v^{\text{sc}})^\top \mathbf 1_n.$ 
Lemma~\ref{lemma_bounds_on_marginals} expresses a lower and upper bounds with respect to $\ell_1$-norm of $\mu^{\text{sc}}$and $\nu^{\text{sc}}$, while Proposition~\ref{proposition_error_in_marginals} gives an upper bound of the marginal errors.

\begin{lemma}
\label{lemma_bounds_on_marginals}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}.
Then one has 
\begin{align*}
 \kappa \norm{\mu_{I_{\varepsilon,\kappa}}^{\text{}}}_1 & + (n -n_b) \Big(\frac{mm_b\varepsilon^2 K_{\min}^2 \min_{j\in J_{\varepsilon, \kappa}} \nu_j}{ (n-n_b)m\kappa \varepsilon^2 K_{\min} + n_b\kappa^2 \max_{i\in I_{\varepsilon,\kappa}}\mu_i } + (m-m_b)\varepsilon^2K_{\min}\Big)\\
 & \leq \norm{\mu^{\text{sc}}}_1  \leq \kappa \norm{\mu_{I_{\varepsilon,\kappa}}^{\text{}}}_1 + (n-n_b) \Big(\frac{m_b\max_{j \in J_{\varepsilon,\kappa}} \nu_j}{n\kappa K_{\min}} + (m-m_b)\varepsilon^2 \Big)
\end{align*}
and 
\begin{align*}
\kappa^{-1} \norm{\nu_{J_{\varepsilon,\kappa}}^{\text{}}}_1 & + (m -m_b) \Big(\frac{nn_b\varepsilon^2\kappa^2K_{\min}^2 \min_{i\in I_{\varepsilon, \kappa}} \mu_j}{ n(m-m_b)\varepsilon^2 \kappa K_{\min} + m_b\kappa \max_{j\in J_{\varepsilon,\kappa}}\nu_j } + (n-n_b)\varepsilon^2K_{\min}\Big)\\
&\norm{\nu^{\text{sc}}}_1 \leq \kappa^{-1} \norm{\nu_{J_{\varepsilon,\kappa}}}_1 + (m-m_b)\Big(\frac{n_b\kappa \max_{i\in I_{\varepsilon,\kappa}}\mu_i}{m K_{\min}} + (n-n_b)\varepsilon^2\Big).
\end{align*}

\end{lemma}
\begin{proposition}
\label{proposition_error_in_marginals}
% Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}. Then, one has 
One has 
\begin{align*}
{\norm{{\mu} -{\mu}^{\text{sc}}}^2_1}&\leq 7\norm{\mu^{\text{sc}}}_1 n_b(\kappa-\log(\kappa)-1)\max_{i} \mu_i + 7\norm{\mu^{\text{sc}}}_1  (n-n_b)\bigg(\frac{m_b\max_{j}\nu_j}{n\kappa K_{\min}} + (m- m_b) \varepsilon^2\\
&\qquad - \min_{i}\mu_i + \max_{i} \mu_i\log\Big(\frac{\kappa(n-n_b+ 1)\max_{i} \mu_i}{m_b K_{\min}\min_{j \in J_{\varepsilon,\kappa}}\nu_j} + \frac{\kappa^2\max_{i} \mu_i}{mm_b\varepsilon^2K_{\min}^2\min_{j \in J_{\varepsilon,\kappa}}\nu_j}\Big)\bigg)
\end{align*}
and 
\begin{align*}
{\norm{{\nu} -{\nu}^{\text{sc}}}^2_1} &\leq 7\norm{\nu^{\text{sc}}}_1m_b(\kappa-\log(\kappa)-1)\max_{j} \nu_j + 7\norm{\nu^{\text{sc}}}_1(m-m_b)\bigg(\frac{n_b\kappa\max_{i}\mu_i}{n K_{\min}} + (n- n_b) \varepsilon^2\\ 
&\qquad - \min_{j}\nu_j + \max_{j} \nu_j\log\Big(\frac{\kappa(m-m_b+ 1)\max_{j} \nu_j}{n_b K_{\min}\min_{i \in I_{\varepsilon,\kappa}}\mu_i} + \frac{\kappa^2\max_{j} \nu_j}{nn_b\varepsilon^2K_{\min}^2\min_{i \in I_{\varepsilon,\kappa}}\mu_i}\Big)\bigg).
\end{align*}
\end{proposition}
Proofs of Lemma~\ref{lemma_bounds_on_marginals} and Proposition~\ref{proposition_error_in_marginals} are presented in supplementary material. 
Proof of Proposition~\ref{proposition_error_in_marginals} is based on a generalization of the Pinsker inequality~\cite{altschulernips17,khalilabid2018}.

>>>>

% We derive in this section 
This section is devoted to establishing theoretical guarantees for the marginal violations of Screenkhorn. 
We first define the screened marginals $\mu^{\text{sc}} = B(u^{\text{sc}}, v^{\text{sc}}) \mathbf 1_m$ and $\nu^{\text{sc}} = B(u^{\text{sc}}, v^{\text{sc}})^\top \mathbf 1_n.$ 
Lemma~\ref{lemma_bounds_on_marginals} expresses a lower and upper bounds with respect to $\ell_1$-norm of $\mu^{\text{sc}}$and $\nu^{\text{sc}}$, while Proposition~\ref{proposition_error_in_marginals} gives an upper bound of marginal errors.

\begin{lemma}
\label{lemma_bounds_on_marginals}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}.
Then one has 
\begin{align*}
 \kappa \norm{\mu_{I_{\varepsilon,\kappa}}^{\text{}}}_1 & + (n -n_b) \Big(\frac{mm_b\varepsilon^2 K_{\min}^2 \min_{j\in J_{\varepsilon, \kappa}} \nu_j}{ (n-n_b)m\kappa \varepsilon^2 K_{\min} + n_b\kappa^2 \max_{i\in I_{\varepsilon,\kappa}}\mu_i } + (m-m_b)\varepsilon^2K_{\min}\Big)\\
 & \leq \norm{\mu^{\text{sc}}}_1  \leq \kappa \norm{\mu_{I_{\varepsilon,\kappa}}^{\text{}}}_1 + (n-n_b) \Big(\frac{m_b\max_{j \in J_{\varepsilon,\kappa}} \nu_j}{n\kappa K_{\min}} + (m-m_b)\varepsilon^2 \Big)
\end{align*}
and 
\begin{align*}
\kappa^{-1} \norm{\nu_{J_{\varepsilon,\kappa}}^{\text{}}}_1 & + (m -m_b) \Big(\frac{nn_b\varepsilon^2\kappa^2K_{\min}^2 \min_{i\in I_{\varepsilon, \kappa}} \mu_j}{ n(m-m_b)\varepsilon^2 \kappa K_{\min} + m_b\kappa \max_{j\in J_{\varepsilon,\kappa}}\nu_j } + (n-n_b)\varepsilon^2K_{\min}\Big)\\
&\norm{\nu^{\text{sc}}}_1 \leq \kappa^{-1} \norm{\nu_{J_{\varepsilon,\kappa}}}_1 + (m-m_b)\Big(\frac{n_b\kappa \max_{i\in I_{\varepsilon,\kappa}}\mu_i}{m K_{\min}} + (n-n_b)\varepsilon^2\Big).
\end{align*}
\end{lemma}
\begin{proposition}
\label{proposition_error_in_marginals}
One has %\norm{\mu^{\text{sc}}}_1
\begin{align*}
{\norm{{\mu} -{\mu}^{\text{sc}}}^2_1}&\leq 7 n_b(\kappa-\log(\kappa)-1)\max_{i} \mu_i + 7 (n-n_b)\bigg(\frac{m_b\max_{j}\nu_j}{n\kappa K_{\min}} + (m- m_b) \varepsilon^2\\
&\qquad - \min_{i}\mu_i + \max_{i} \mu_i\log\Big(\frac{\kappa(n-n_b+ 1)\max_{i} \mu_i}{m_b K_{\min}\min_{j \in J_{\varepsilon,\kappa}}\nu_j} + \frac{\kappa^2\max_{i} \mu_i}{mm_b\varepsilon^2K_{\min}^2\min_{j \in J_{\varepsilon,\kappa}}\nu_j}\Big)\bigg)
\end{align*}
and 
\begin{align*}
{\norm{{\nu} -{\nu}^{\text{sc}}}^2_1} &\leq 7 m_b(\kappa-\log(\kappa)-1)\max_{j} \nu_j + 7(m-m_b)\bigg(\frac{n_b\kappa\max_{i}\mu_i}{n K_{\min}} + (n- n_b) \varepsilon^2\\ 
&\qquad - \min_{j}\nu_j + \max_{j} \nu_j\log\Big(\frac{\kappa(m-m_b+ 1)\max_{j} \nu_j}{n_b K_{\min}\min_{i \in I_{\varepsilon,\kappa}}\mu_i} + \frac{\kappa^2\max_{j} \nu_j}{nn_b\varepsilon^2K_{\min}^2\min_{i \in I_{\varepsilon,\kappa}}\mu_i}\Big)\bigg).
\end{align*}
\end{proposition}
Proofs of Lemma~\ref{lemma_bounds_on_marginals} and Proposition~\ref{proposition_error_in_marginals} are presented in supplementary material. 
Proof of Proposition~\ref{proposition_error_in_marginals} is based on a generalization of the Pinsker inequality~\cite{altschulernips17,khalilabid2018}.
